\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-21 paper.qmd
Find one of most appropriate model for PBMC

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC_"
)

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 132-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with UMAP, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters."
#| label: fig-model-pbmc-author
#| fig-pos: H

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = umap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_umap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = tsne_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_tsne,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = phate_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_phate,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = trimap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_trimap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.1)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = pacmap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_pacmap,
    q = 0.1, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

error_plot_pbmc <- ggplot(error_pbmc, 
                              aes(x = b_non_empty, 
                                  y = log(Error), 
                                  colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  geom_vline(xintercept = 124, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")+
    geom_segment(data = arrow_df_pbmc, aes(
      x = x,  
      y = y,  
      xend = xend, 
      yend = yend), 
      arrow = arrow(length = unit(0.01, "npc"),
        ends = "first"), 
        color = "black") +
    inset_element(p = tsne_5_img, 
                  left = 1.7, 
                  bottom = 0.16, 
                  right = 0, 
                  top = 0.31) +                   
    inset_element(p = phate_5_img, 
                  left = 0.6, 
                  bottom = 0.5, 
                  right = 0, 
                  top = 0.65) +                   
    inset_element(p = umap_img, 
                  left = 1.3, 
                  bottom = 0.3, 
                  right = 0, 
                  top = 0.5) +
    inset_element(p = tsne_30_img, 
                  left = 0.6, 
                  bottom = 0.03, 
                  right = 0, 
                  top = 0.23) +
  theme(#aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90),
        legend.position = "none",
        legend.text = element_text(size = 7))

@@

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to $124$, the lowest error occurred with the parameters perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 70%

@@

<!--Fit the best model and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_pbmc <- 16
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC_"
)

@@

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(
  distance_edges = distance_pbmc,
  distance_col = "distance")
#benchmark_pbmc <- 0.1

@@

<!--bin1 = 15, bin2 = 20, b = 300, non_empty = 137-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with tSNE, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters, but most low $p\\text{-}D$ model errors present in the large cluster."
#| label: fig-model-pbmc
#| fig-pos: H
#| out-height: 30%

>>>>

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.05,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC_"
)

@@

<!--bin1 = 30, bin2 = 29, b = 870, non_empty = 135-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with UMAP, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters."
#| label: fig-model-pbmc-author
#| fig-pos: H

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.05)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = umap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_umap,
    q = 0.05, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.05)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = tsne_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_tsne,
    q = 0.05, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.05)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = phate_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_phate,
    q = 0.05, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.05)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = trimap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_trimap,
    q = 0.05, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.05)$bin2

  pbmc_model <- fit_highd_model(
    training_data = training_data_pbmc,
    emb_df = pacmap_pbmc_scaled,
    bin1 = xbins,
    r2 = r2_pacmap,
    q = 0.05, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "PC_"
  )

@@

error_plot_pbmc <- ggplot(error_pbmc, 
                              aes(x = b_non_empty, 
                                  y = log(Error), 
                                  colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  geom_vline(xintercept = 136, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")+
    geom_segment(data = arrow_df_pbmc, aes(
      x = x,  
      y = y,  
      xend = xend, 
      yend = yend), 
      arrow = arrow(length = unit(0.01, "npc"),
        ends = "first"), 
        color = "black") +
    inset_element(p = tsne_5_img, 
                  left = 1.7, 
                  bottom = 0.16, 
                  right = 0, 
                  top = 0.31) +                   
    inset_element(p = phate_5_img, 
                  left = 0.6, 
                  bottom = 0.5, 
                  right = 0, 
                  top = 0.65) +                   
    inset_element(p = umap_img, 
                  left = 1.3, 
                  bottom = 0.3, 
                  right = 0, 
                  top = 0.5) +
    inset_element(p = tsne_30_img, 
                  left = 0.6, 
                  bottom = 0.03, 
                  right = 0, 
                  top = 0.23) +
  theme(#aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90),
        legend.position = "none",
        legend.text = element_text(size = 7))

@@

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to $136$, the lowest error occurred with the parameters perplexity: $30$."
#| label: fig-pbmc-abserror
#| fig-pos: H
#| out-width: 70%

@@

<!--Fit the best model and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_pbmc <- 15
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.05,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC_"
)

@@

@@

<!--bin1 = 15, bin2 = 28, b = 420, non_empty = 136-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in $2\\text{-}D$ with tSNE, and (b) $p\\text{-}D$ model error in $2\\text{-}D$. The $2\\text{-}D$ model shows three well-separated distant clusters. The $p\\text{-}D$ model errors are distributed along clusters, but most low $p\\text{-}D$ model errors present in the large cluster."
#| label: fig-model-pbmc
#| fig-pos: H
#| out-height: 30%

### 2024-07-21 paper.tex
Added 2D model overlaid on data (MNIST data)
Find one of most appropriate model for PBMC

\caption{\label{fig-pbmc-abserror}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to \(124\), the
lowest error occurred with the parameters perplexity: \(30\).}

@@

\begin{minipage}{0.33\linewidth}

@@

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

>>>>

\caption{\label{fig-pbmc-abserror}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to \(136\), the
lowest error occurred with the parameters perplexity: \(30\).}

@@

\begin{minipage}{0.25\linewidth}

\includegraphics{figures/mnist/mnist_model_2d.png}

\subcaption{\label{fig-mnist1-model}}

\end{minipage}%
%
\begin{minipage}{0.25\linewidth}

@@

\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
